Twenty years of dynamic occupancy models: a review of applications and a look towards the future

### 2024-08-29 Manuscript.qmd
shuffled directory for Manuscript

abstract: | 
  Describing patterns of species occupancy across landscapes and throughout time is a key objective of much ecological research. Nonetheless, reliably estimating occupancy can be a challenge, particularly when common issues such as imperfect detection and shifting distributions are at play. The dynamic, multi-season occupancy model (DOM) offers a framework for occupancy estimation which explicitly accounts for observation error, while also incorporating additional mechanistic process by explicitly estimating colonisation and extinction rates. In the two decades since their introduction the DOM has become a popular tool for a popular tool for describing patterns in occupancy, exploring the environmental drivers of species occurence, and, more recently, for predicting species occupancy to new locations and to future conditions.

  In this review we examine how ecologists have applied the DOM in its first two decades of use. Based on a sample of articles fitting DOMs to field ecological data, we explore the systems in which these models have been used, how data was collected for them, and how authors have built their models to achieve their objectives. Our findings indicate that the DOM is an particularly flexible tool readily applied to a range of diverse taxa from around the globe, capable of being used to estimate occupancy at study sites ranging from local to continental scales over time periods from months to decades. The DOM framework is also amenable to extension, further broadening their utility to address a wide range of ecological questions.

  A key aim of this review is to explore how users of DOMs have incorporated covariates into their models to describe variation in occupancy through space and time. Broadly speaking, model complexity in DOMs tends to be low, with relatively few covariates considered for inclusion and most relationships represented as simple linear terms. Considerable variation also exists in the procedures used for covariate selection, and limited research has been conducted on how these choices may influence model performance and inference. Furthermore, only a fraction of articles appear to evaluate models (either by goodness-of-fit testing or prediction validation). Likewise, little guidance exists on how to approach this task in DOMs. These uncertainties in the modelling process should be key priorities for future research on DOMs given their increasing use in applied ecological research.
---

@@

The site occupancy models first introduced by @mackenzie2002 offer a powerful framework for modelling presence/absence data while accounting for each of these challenges. Originating in the mark-recapture literature, occupancy models use multiple resurveys of sites to estimate detection probabilities and correct for bias in estimates of site occupancy. Originally a static model, @mackenzie2003 extended this model for use in multiple time-steps by explicitly describing the process of changing occupancy via colonisation and extinction, relaxing assumptions of equilibrium and allowing description of patterns of site occupancy through time. This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly incorporating the key processes which drive occupancy while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic data required by more process-explicit models [@briscoe2019]. These features make the DOM an important tool, with uses for tasks including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@kéry2013; @briscoe2021].

@@

The DOM unites two processes: the ecological process of site occupancy describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed given that a site is occupied (@fig-modelform). Under this parameterisation, sites exist in either an occupied or unoccupied state. In the first time-step, occupancy state at each site is determined by the probability of initial occupancy. In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation and extinction. In the observation component of the DOM, the model explicitly accounts for imperfect detection. At occupied sites, the detection probability describes whether or not the species is observed during a survey; under the DOM's original parameterisation, it is assumed that false-positive detections at unoccupied sites never occur.

![The form of the dynamic occupancy model as described by @mackenzie2003. The ecological process submodel describes changes in occupancy over time, where occupancy in the first season is given by the probability of initial occupancy and occupancy in ensuing seasons is governed by colonisation and extinction probabilities. The observational process submodel describes detectability, where the probability of detecting a species given that a site is occupied is given by the detection probability. At sites which are truly unoccupied, is it assumed that the species is never detected.](Figures/ModelForm.png){#fig-modelform}

To disentangle the ecological and observational processes the DOM requires a hierarchical sampling design as depicted in @fig-surveys. Under this design, observations at a site occur during distinct, time-bound seasons within which it is assumed site occupancy does not change (that is, sites are closed to changes in occupancy). In each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy. Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@mackenzie2005].

![The sampling design of the standard dynamic occupancy model. During seasons, also called primary occasions, sites are considered closed to changes in true occupancy state; occupancy state may only change between seasons. Within each season, multiple observations ('secondary occasions') are conducted to record the observed presence or absence of the species at each site. These multiple observations may be recorded in many ways: sites can be revisited multiple times within a season, surveys can be conducted at multiple points within a larger site, multiple observers can conduct surveys contemporaneously, or the time elapsed prior to a detection occurring can be recorded. It is not necessary for the same number of observations to occur in each year or each site, allowing for greater flexibility in data inputs.](Figures/SurveyDesign.png){#fig-surveys fig-align="center"}

This description of the DOM model form and sampling design are based on the original form of the dynamic occupancy model as prescribed in @mackenzie2003. Numerous model extensions and alternative formulations are also available, including implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models [@dorazio2010]. For a comprehensive discussion of the most common extensions and their applications see @bailey2014, as well as @devarajan2020 for a more detailed review of multi-species occupancy models.

DOMs make a number of assumptions requiring careful consideration by model users, which we outline here as our review interrogates related aspects of model building.

I.  **False positive detections do not occur**. While this assumption can be safely met in many studies, it is not necessarily guaranteed when working with more cryptic species or less reliable survey methods. @mcclintock2010 and @miller2015 comment on the bias induced when false positives occur and are not accounted for, highlighting the need for authors to consider how certain their detections truly are. Significantly, even genuine detections of a species can be considered 'false positives' when they do not represent the intended definition of occupancy, such as detections of transient individuals when the intent is to estimate breeding occupancy [@berigan2019]. Where this assumption can not reasonably met, aforementioned extensions designed to account for false positive error should be considered.

@@

III. **Heterogeneity in occupancy and detection is accounted for.** As with any approach for modelling species occurrence, it is assumed that DOMs appropriately capture variation in occupancy patterns and species detectability across the study system. Generally, this is achieved by allowing model parameters to vary with respect to covariates representing the environmental factors which may be expected to influence these parameters. An important element of this assumption is that the likelihood of detection of a species depends not only on the observability of the species, but also on factors like habitat suitability which influence species abundance and activity [@guillera-arroita2017]. While no model will fully account for the complexity inherent in patterns of species occupancy and detection, failure to capture key drivers is likely to introduce bias and confound inference made from model estimates. Compared to the first two assumptions mentioned, this aspect of DOMs has been less thoroughly discussed and comparatively little is known how this latent heterogeneity can influence model estimates.

@@

To assess how DOMs have been applied in the years since their introduction we gathered a representative sample of articles fitting them to field ecological data. A pool of candidate articles was generated using two queries on Web of Science: the first of these included all articles from 2004-2023 which cite @mackenzie2003; the second query searched articles in the same time-span matching "dynamic occupancy model\*", "multi-season occupancy model\*", and "occupancy dynamic\*", articles including each of "occupancy", "colonization", "extinction/persistence", and "detection", and articles with the term "occupancy" located near "dynamic" in the title, key terms, or abstract. As we were interested in how DOMs use has changed through time, we divided these articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024. From each of these strata we randomly selected 20 articles for inclusion in the review. Articles which did not meet inclusion criteria were replaced from within their own strata.

@@

We recorded details on the type of organisms (bird, mammal, etc.) modelled in each article, and the means by which multiple species were modelled where applicable. Taxa were denoted as threatened either when they are listed on the IUCN Red List of Species or when authors otherwise indicate that they are threatened. This deference to authors' representation of conservation status was made to account for sub-species which lack listings or species which are of more local concern. Study locality and size was documented; the size of the study area being defined as the intended area of inference which contained all site — this was measured to the order of magnitude to account for uncertainty in reporting.

We were particularly interested in how authors navigated the modelling process, from covariate selection through model evaluation. To this end, we recorded all covariates considered in each study regardless of whether they were or were not included in final models. Key traits of each covariate were recorded including their general category, whether they were directly observed or remotely sensed, whether they were static or varied between seasons, and how they were presented in the model (as a linear term, a polynomial term, or as part of an interaction with another covariate [@james2021]). Model selection procedures were also sorted into non-exclusive categories including *A priori*, where only one model was considered; *candidate suite*, where a predefined set of models was considered; *sequential,* where covariates were selected parameter-by-parameter; and *simple precursors*, where selection was preceded by another simpler model implementation. Finally, for each model we documented whether goodness-of-fit was tested and reported, as well as whether model performance was assessed by validation with either in or out-of-sample data. For the full spreadsheet of data collected and further details on categorisation, see **Appendix I.**

>>>>

abstract: | 
  Describing patterns of species occupancy across landscapes and throughout time is a key objective of much ecological research. Nonetheless, reliably estimating occupancy can be a challenge, particularly when common issues such as imperfect detection and shifting distributions are at play. The dynamic, multi-season occupancy model (DOM) offers a framework for occupancy estimation which explicitly accounts for observation error, while also incorporating additional mechanistic process by explicitly estimating colonisation and extinction rates. In the two decades since its introduction, the DOM has become a popular tool for describing patterns in occupancy, exploring the environmental drivers of species occurence, and, more recently, for predicting species occupancy to new locations and to future conditions.

  In this review we examine how ecologists have applied the DOM in its first two decades of use. Based on a sample of articles fitting DOMs to field ecological data, we explore the systems in which these models have been used, how data were collected for them, and how authors have built their models to achieve their objectives. Our findings indicate that the DOM is an particularly flexible tool readily applied to a range of diverse taxa from around the globe, capable of being used to estimate occupancy at study sites ranging from local to continental scales over time periods from months to decades. The DOM framework is also amenable to extension, further broadening their utility to address a wide range of ecological questions.

  A key aim of this review is to explore how users of DOMs have incorporated covariates into their models to describe variation in occupancy through space and time. Broadly speaking, model complexity in DOMs tends to be low, with relatively few covariates considered for inclusion and most relationships represented as simple linear terms. Considerable variation also exists in the procedures used for covariate selection, and limited research has been conducted on how these choices may influence model performance and inference. Furthermore, only a fraction of articles appear to evaluate models (either by goodness-of-fit testing or prediction validation). Likewise, little guidance exists on how to approach this task in DOMs. These uncertainties in the modelling process should be key priorities for future research on DOMs given their increasing use in applied ecological research.
bibliography: references.bib
---

@@

The site occupancy models first introduced by @mackenzie2002 set the foundation for a powerful framework for modelling presence/absence data while accounting for each of these challenges [@guillera-arroita2017]. Drawing on principles from the mark-recapture literature, occupancy models use multiple resurveys of sites to estimate detection probabilities and correct for bias in estimates of site occupancy. Originally a static model, @mackenzie2003 extended this model for use in multiple time-steps by explicitly describing the process of changing occupancy via colonisation and extinction, relaxing assumptions of equilibrium and allowing description of patterns of site occupancy through time. This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly describing the key processes which drive occupancy dynamics while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic data required by more process-explicit models [@briscoe2019]. These features make the DOM an important tool, with uses for tasks including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@kéry2013; @briscoe2021].

@@

The DOM structure encompasses two processes: the ecological process of site occupancy dynamics describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed at a site where it is present (@fig-modelform). Sites exist in either an occupied or unoccupied state. In the first time step, occupancy state at each site is determined by the probability of initial occupancy. In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation and local extinction. In the observation component of the DOM, the model explicitly accounts for imperfect detection. At occupied sites, the detection probability describes whether or not the species is observed during a survey; under the DOM's original parameterisation, it is assumed that false-positive detections at unoccupied sites never occur.

![The form of the dynamic occupancy model as described by @mackenzie2003. The ecological process submodel describes changes in occupancy over time, where occupancy in the first season is given by the probability of initial occupancy and occupancy in ensuing seasons is governed by colonisation and extinction probabilities. The observational process submodel describes detectability, where the probability of detecting a species given that a site is occupied is given by the detection probability. At sites which are truly unoccupied, it is assumed that the species is never detected (i.e., no false positives).](Figures/ModelForm.png){#fig-modelform}

To disentangle the ecological and observational processes the DOM requires a hierarchical sampling design as depicted in @fig-surveys. Under this design, observations at a site occur during distinct, time-bound seasons within which it is assumed site occupancy does not change (that is, sites are closed to changes in occupancy). In each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy. Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@mackenzie2005]. It is important to note that it is not necessary for the same number of observations to occur in each year or for each site, and that the DOM accommodates for missing data; this allows for considerably greater flexibility in data inputs.

![The sampling design of the standard dynamic occupancy model. During seasons, also called primary occasions, sites are considered closed to changes in true occupancy state; occupancy state may only change between seasons. Within each season, multiple observations ('secondary occasions') are conducted to record the observed presence or absence of the species at each site. These multiple observations may be recorded in many ways: sites can be revisited multiple times within a season, surveys can be conducted at multiple points within a larger site, multiple observers can conduct surveys contemporaneously, or the time elapsed prior to a detection occurring can be recorded. Note that it is not necessary for each site or season to have the same number of observations, and that missing data can be accommodated.](Figures/SurveyDesign.png){#fig-surveys fig-align="center"}

DOMs make a number of key assumptions requiring careful consideration by model users, which we outline here as our review interrogates related aspects of model building.

I.  **False positive detections do not occur**. While this assumption can be safely met in many studies, it is not necessarily guaranteed when working with more cryptic species or less reliable survey methods. @mcclintock2010 and @miller2015 comment on the bias induced when false positives occur and are not accounted for, highlighting the need for authors to consider how certain their detections truly are. Significantly, even genuine detections of a species can be considered 'false positives' when they do not represent the intended definition of occupancy, such as detections of transient individuals when the intent is to estimate breeding occupancy [@berigan2019]. Where this assumption can not reasonably met, model extensions designed to account for false positive error should be considered.

@@

III. **Heterogeneity in occupancy and detection is accounted for.** As with any approach for modelling species occurrence, it is assumed that DOMs appropriately capture variation in occupancy patterns and species detectability across the study system. Generally, this is achieved by allowing model parameters to vary with respect to covariates representing the environmental factors which may be expected to influence these parameters. An important element of this assumption is that the likelihood of detection of a species depends not only on the observability of the species, but also on factors like habitat suitability which influence species abundance and activity [@guillera-arroita2017]. While no model will fully account for the complexity inherent in patterns of species occupancy and detection, failure to capture key drivers is likely to introduce bias and confound inference made from model estimates. Compared to the first two assumptions mentioned, this aspect of DOMs has been less thoroughly discussed and comparatively little is known about how this latent heterogeneity can influence model performance.

Since its original description in @mackenzie2003, the DOM has been further developed with numerous model extensions and alternative formulations including implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models [@dorazio2010]. For a comprehensive discussion of the most common extensions and their applications see @bailey2014 and @guillera-arroita2017, as well as @devarajan2020 for a more detailed review of multi-species occupancy models.

@@

To assess how DOMs have been applied in the years since their introduction we gathered a representative sample of articles fitting them to field ecological data. A pool of candidate articles was generated using two queries on Web of Science. The first of these included all articles from 2004-2023 which cite @mackenzie2003. To ensure that relevant articles which did not directly cite @mackenzie2003 were captured, a second query searched articles in the same time-span matching the terms "*dynamic occupancy model\**", "*multi-season occupancy model\**", or "*occupancy dynamic\**"; articles including each of "*occupancy*", "*colonization*", "*extinction/persistence*", and "*detection*"; and articles with the term "*occupancy*" located near "*dynamic*" in the title, key terms, or abstract. As we were interested in how DOMs use has changed through time, we divided all articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024. From each of these strata we randomly selected 20 articles for inclusion in the review. Articles which did not meet inclusion criteria were replaced from within their own strata.

@@

We recorded details on the type of organisms (bird, mammal, etc.) modelled in each article, and the means by which multiple species were modelled where applicable. Taxa were denoted as threatened either when they are listed on the IUCN Red List of Species or when authors indicate that they are threatened. This deference to authors' representation of conservation status was made to account for sub-species which lack listings or species which are of more local concern. Study locality and size was documented; the size of the study area being defined as the intended area of inference which contained all sites — this was measured to the order of magnitude to account for uncertainty in reporting.

We were particularly interested in how authors navigated the modelling process, from covariate selection through model evaluation. To this end, we recorded all covariates considered in each study regardless of whether they were or were not included in final models. Key traits of each covariate were recorded including their general category, whether they were directly observed or remotely sensed, whether they were static or varied between seasons, and how they were presented in the model (as a linear term, a polynomial term, or as part of an interaction with another covariate [@james2021]). Model selection procedures were also sorted into non-exclusive categories including *A priori*, where only one model was considered; *candidate suite*, where a predefined set of models was considered; *sequential,* where covariates were selected parameter-by-parameter; and *simple precursors*, where selection was preceded by another simpler model implementation. Finally, for each model we documented whether goodness-of-fit was tested and reported, as well as whether model performance was assessed by validation with either in or out-of-sample data. For the full spreadsheet of data collected and further details on categorisation, see supplementary material.

### 2024-08-29 _manuscript/Manuscript.qmd
shuffled directory for Manuscript

The DOM unites two processes: the ecological process of site occupancy describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed given that a site is occupied (@fig-modelform). Under this parameterisation, sites exist in either an occupied or unoccupied state. In the first time-step, occupancy state at each site is determined by the probability of initial occupancy. In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation and extinction. In the observation component of the DOM, the model explicitly accounts for imperfect detection. At occupied sites, the detection probability describes whether or not the species is observed during a survey; under the DOM's original parameterisation, it is assumed that false-positive detections at unoccupied sites never occur.

>>>>

The DOM structure encompasses two processes: the ecological process of site occupancy dynamics describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed at a site where it is present (@fig-modelform). Sites exist in either an occupied or unoccupied state. In the first time step, occupancy state at each site is determined by the probability of initial occupancy. In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation and local extinction. In the observation component of the DOM, the model explicitly accounts for imperfect detection. At occupied sites, the detection probability describes whether or not the species is observed during a survey; under the DOM's original parameterisation, it is assumed that false-positive detections at unoccupied sites never occur.
